---
title: "Project 7: Difference-in-Differences and Synthetic Control"
output:
  html_document: default
  pdf_document: default
---


```{r}
# Install and load packages 
if (!require("pacman")) install.packages("pacman")

devtools::install_github("ebenmichael/augsynth")

pacman::p_load(# Tidyverse packages including dplyr and ggplot2 
               tidyverse,
               ggthemes,
               augsynth,
               gsynth)

# set seed
set.seed(44)

# load data 
medicaid_expansion <- read_csv('C:/Users/lande/Downloads/medicaid_expansion.csv')


```

# Introduction

For this project, you will explore the question of whether the Affordable Care Act increased health insurance coverage (or conversely, decreased the number of people who are uninsured). The ACA was passed in March 2010, but several of its provisions were phased in over a few years. The ACA instituted the "individual mandate" which required that all Americans must carry health insurance, or else suffer a tax penalty. There are four mechanisms for how the ACA aims to reduce the uninsured population:

- Require companies with more than 50 employees to provide health insurance.
- Build state-run healthcare markets ("exchanges") for individuals to purchase health insurance.
- Provide subsidies to middle income individuals and families who do not qualify for employer based coverage.
- Expand Medicaid to require that states grant eligibility to all citizens and legal residents earning up to 138\% of the federal poverty line. The federal government would initially pay 100\% of the costs of this expansion, and over a period of 5 years the burden would shift so the federal government would pay 90\% and the states would pay 10\%.

In 2012, the Supreme Court heard the landmark case NFIB v. Sebelius, which principally challenged the constitutionality of the law under the theory that Congress could not institute an individual mandate. The Supreme Court ultimately upheld the individual mandate under Congress's taxation power, but struck down the requirement that states must expand Medicaid as impermissible subordination of the states to the federal government. Subsequently, several states refused to expand Medicaid when the program began on January 1, 2014. This refusal created the "Medicaid coverage gap" where there are indivudals who earn too much to qualify for Medicaid under the old standards, but too little to qualify for the ACA subsidies targeted at middle-income individuals.

States that refused to expand Medicaid principally cited the cost as the primary factor. Critics pointed out however, that the decision not to expand primarily broke down along partisan lines. In the years since the initial expansion, several states have opted into the program, either because of a change in the governing party, or because voters directly approved expansion via a ballot initiative.

You will explore the question of whether Medicaid expansion reduced the uninsured population in the U.S. in the 7 years since it went into effect. To address this question, you will use difference-in-differences estimation, and synthetic control.

# Data

The dataset you will work with has been assembled from a few different sources about Medicaid. The key variables are:

- **State**: Full name of state
- **Medicaid Expansion Adoption**: Date that the state adopted the Medicaid expansion, if it did so.
- **Year**: Year of observation.
- **Uninsured rate**: State uninsured rate in that year.

# Exploratory Data Analysis

Create plots and provide 1-2 sentence analyses to answer the following questions:

- Which states had the highest uninsured rates prior to 2014? The lowest?

#The states with the highest uninsured rates prior to 2014 were primarily in the South and Southwest, with Texas likely at the top, followed by states like Florida and Nevada. States with the lowest uninsured rates were concentrated in the Northeast, with Massachusetts showing the lowest.

- Which states were home to most uninsured Americans prior to 2014? How about in the last year in the data set? **Note**: 2010 state population is provided as a variable to answer this question. In an actual study you would likely use population estimates over time, but to simplify you can assume these numbers stay about the same.

#Prior to 2014, the states with the most uninsured people were those with large populations and high uninsured rates, with Texas, California, and Florida likely topping the list. By the most recent year in the dataset, we can see a shift in the distribution, especially in states that expanded Medicaid, while non-expansion states like Texas likely maintained higher absolute numbers of uninsured individuals.


```{r}
medicaid_expansion <- read.csv('C:/Users/lande/Downloads/medicaid_expansion.csv')

pre_2014_uninsured <- medicaid_expansion %>%
  filter(year < 2014) %>%
  group_by(State) %>%
  summarize(Avg_uninsured_rate = mean(`uninsured_rate`, na.rm = TRUE)) %>%
  arrange(desc(Avg_uninsured_rate))

# highest and lowest uninsured rates
```
# Plot top 10 highest uninsured rates pre-2014
```{r}
p1 <- pre_2014_uninsured %>%
  slice_max(order_by = Avg_uninsured_rate, n = 10) %>%
  ggplot(aes(x = reorder(State, Avg_uninsured_rate), y = Avg_uninsured_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 States with Highest Uninsured Rates (Pre-2014)",
       x = "State", y = "Average uninsured rate (%)") +
  theme_minimal()

p1


```
```{r}
# Plot top 10 lowest uninsured rates pre-2014

p2 <- pre_2014_uninsured %>%
  slice_min(order_by = Avg_uninsured_rate, n = 10) %>%
  ggplot(aes(x = reorder(State, -Avg_uninsured_rate), y = Avg_uninsured_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 States with Lowest Uninsured Rates (Pre-2014)",
       x = "State", y = "Average uninsured rate (%)") +
  theme_minimal()

p2


```
# most uninsured Americans
```{r}
medicaid_expansion <- medicaid_expansion %>%
  mutate(uninsured_Count = `uninsured_rate` * population / 100)
```
```{r}
# Most uninsured people prior to 2014
pre_2014_uninsured_count <- medicaid_expansion %>%
  filter(year == 2013) %>%  # Using 2013 as the last year before expansion
  arrange(desc(uninsured_Count)) %>%
  select(State, uninsured_Count, population, `uninsured_rate`)

```
```{r}
# Most uninsured people in the latest year available
latest_year <- max(medicaid_expansion$year)
latest_uninsured_count <- medicaid_expansion %>%
  filter(year == 2020) %>%
  arrange(desc(uninsured_Count)) %>%
  select(State, uninsured_Count, population, `uninsured_rate`)

```
```{r}
library(scales)

p3 <- pre_2014_uninsured_count %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(State, uninsured_Count), y = uninsured_Count / 1000000)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(title = "States with Most Uninsured People in 2013",
       x = "State", y = "Number of Uninsured (Millions)") +
  theme_minimal() +
  scale_y_continuous(labels = comma)

p4 <- latest_uninsured_count %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(State, uninsured_Count), y = uninsured_Count / 1000000)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(title = paste("States with Most Uninsured People in", latest_year),
       x = "State", y = "Number of Uninsured (Millions)") +
  theme_minimal() +
  scale_y_continuous(labels = comma)

# Display the plot
p3


p4


```



# Difference-in-Differences Estimation

## Estimate Model

Do the following:

- Choose a state that adopted the Medicaid expansion on January 1, 2014 and a state that did not. **Hint**: Do not pick Massachusetts as it passed a universal healthcare law in 2006, and also avoid picking a state that adopted the Medicaid expansion between 2014 and 2015.
- Assess the parallel trends assumption for your choices using a plot. If you are not satisfied that the assumption has been met, pick another state and try again (but detail the states you tried).

```{r}
# Parallel Trends plot
did_states <- c("Arizona", "Texas")
did_data <- medicaid_expansion %>%
  filter(State %in% did_states)
```
```{r}
did_plot <- ggplot(did_data, aes(x = year, y = `uninsured_rate`, color = State, group = State)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "black") +
  annotate("text", x = 2014, y = max(did_data$`uninsured_rate`, na.rm = TRUE), 
           label = "Medicaid Expansion", hjust = -0.1) +
  labs(title = "Uninsured Rates: Arizona (Expanded) vs Texas (Did Not Expand)",
       x = "Year", y = "Uninsured Rate (%)") +
  theme_minimal()

print(did_plot)

```

- Estimates a difference-in-differences estimate of the effect of the Medicaid expansion on the uninsured share of the population. You may follow the lab example where we estimate the differences in one pre-treatment and one post-treatment period, or take an average of the pre-treatment and post-treatment outcomes

```{r}
# Calculate DiD estimate
pre_data <- did_data %>% 
  filter(year < 2014) %>%
  group_by(State) %>%
  summarize(Pre_Uninsured = mean(`uninsured_rate`, na.rm = TRUE))

post_data <- did_data %>% 
  filter(year >= 2014) %>%
  group_by(State) %>%
  summarize(Post_Uninsured = mean(`uninsured_rate`, na.rm = TRUE))

did_result <- pre_data %>%
  inner_join(post_data, by = "State") %>%
  mutate(
    Difference = Post_Uninsured - Pre_Uninsured,
    Treatment = State == "Arizona"
  )
```
# Difference-in-Differences estimation
```{r}
# Calculate the DiD estimate
treatment_diff <- did_result$Difference[did_result$Treatment]
control_diff <- did_result$Difference[!did_result$Treatment]
did_estimate <- treatment_diff - control_diff

# Print DiD results
print(did_result)
cat("DiD Estimate:", round(did_estimate, 2), 
    "percentage points reduction in uninsured rate attributable to Medicaid expansion\n")

```
## Discussion Questions

- Card/Krueger's original piece utilized the fact that towns on either side of the Delaware river are likely to be quite similar to one another in terms of demographics, economics, etc. Why is that intuition harder to replicate with this data?
- **Answer**:The intuition from Card/Krueger is harder to replicate with state-level Medicaid data because states are much larger, more heterogeneous units than adjacent towns. Unlike towns on opposite sides of a river that share economic conditions, labor markets, and demographics, states differ substantially in population characteristics, economic conditions, political environments, and existing healthcare infrastructures. Additionally, states made Medicaid expansion decisions based on political factors rather than randomly, creating potential selection bias that wasn't present in the natural geographic boundary used in the Delaware river case.


- What are the strengths and weaknesses of using the parallel trends assumption in difference-in-differences estimates?
- **Answer**:The parallel trends assumption's strength is that it allows for non-random treatment assignment by accounting for time-invariant differences between treatment and control groups, making causal inference possible when randomization isn't feasible. However, its weaknesses include,it's untestable in the post-treatment period; it's vulnerable to time-varying confounders that affect groups differently; pre-treatment parallel trends don't guarantee the counterfactual would have continued in parallel; and it's sensitive to the selection of comparison units. 

# Synthetic Control

Estimate Synthetic Control

Although several states did not expand Medicaid on January 1, 2014, many did later on. In some cases, a Democratic governor was elected and pushed for a state budget that included the Medicaid expansion, whereas in others voters approved expansion via a ballot initiative. The 2018 election was a watershed moment where several Republican-leaning states elected Democratic governors and approved Medicaid expansion. In cases with a ballot initiative, the state legislature and governor still must implement the results via legislation. For instance, Idaho voters approved a Medicaid expansion in the 2018 election, but it was not implemented in the state budget until late 2019, with enrollment beginning in 2020.

Do the following:

- Choose a state that adopted the Medicaid expansion after January 1, 2014. Construct a non-augmented synthetic control and plot the results (both pre-treatment fit and post-treatment differences). Also report the average ATT and L2 imbalance.

```{r}
# non-augmented synthetic control
synth_data <- medicaid_expansion %>%
  # Remove states that expanded on Jan 1, 2014 from donor pool
  mutate(
    Expansion_2014_Jan = !is.na(Date_Adopted) & 
      year(Date_Adopted) == 2014 & 
      month(Date_Adopted) == 1 & 
      day(Date_Adopted) == 1
  )

```
```{r}
synth_wide <- synth_data %>%
  select(State, year, `uninsured_rate`, Date_Adopted) %>%
  # Create treatment indicator for Michigan in April 2014
  mutate(
    Treated = (State == "Michigan"),
    Treatment_Year = ifelse(State == "Michigan", 2014.25, Inf)  # April 2014 = 2014.25
  )

# Create donor pool - states that never expanded Medicaid
donor_states <- synth_data %>%
  filter(is.na(Date_Adopted)) %>%
  pull(State)

```
```{r}
# Regular synthetic control for Michigan
install.packages("remotes")
remotes::install_github("ebenmichael/augsynth")
```
```{r}
library(augsynth)

mi_synth <- augsynth(
  `uninsured_rate` ~ Treated, 
  State, 
  year, 
  synth_wide %>% filter(State == "Michigan" | State %in% donor_states),
  t_int = 2014.25, # April 2014
  progfunc = "None"  # Standard synthetic control without augmentation
)

```
```{r}
# Plot results
plot(mi_synth)

# Get ATT and L2 imbalance
mi_att <- summary(mi_synth)
print(mi_att)
```

- Re-run the same analysis but this time use an augmentation (default choices are Ridge, Matrix Completion, and GSynth). Create the same plot and report the average ATT and L2 imbalance.

```{r}
# augmented synthetic control
mi_augsynth_ridge <- augsynth(
  `uninsured_rate` ~ Treated, 
  State, 
  year, 
  synth_wide %>% filter(State == "Michigan" | State %in% donor_states),
  t_int = 2014.25,
  progfunc = "Ridge"
)

```
```{r}
# Plot results for augmented synthetic control
plot(mi_augsynth_ridge)

# Get ATT and L2 imbalance for augmented version
mi_aug_att <- summary(mi_augsynth_ridge)
print(mi_aug_att)

```
- Plot barplots to visualize the weights of the donors.
```{r}
str(mi_synth$weights)
```

```{r}
# For standard synthetic control
# barplots of weights
weights_df %>%
  filter(Weight > 0.01) %>%
  arrange(desc(Weight))

weights_df %>%
  filter(Weight > 0.01) %>%
  arrange(Weight) %>%
  ggplot(aes(x = reorder(State, Weight), y = Weight)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Synthetic Control Weights for Michigan",
       x = "Donor State", y = "Weight") +
  theme_minimal()
```
# For augmented synthetic control

```{r}
aug_weights_df <- data.frame(
  State = rownames(aug_weights),
  Weight = as.numeric(aug_weights)
) %>%
  filter(abs(Weight) > 0.01) %>%
  arrange(desc(Weight))

```
```{r}
# Plot augmented weights
p6 <- ggplot(aug_weights_df, aes(x = reorder(State, Weight), y = Weight, 
                                fill = Weight > 0)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("darkred", "darkgreen"), 
                    labels = c("Negative Weight", "Positive Weight")) +
  coord_flip() +
  labs(title = "Augmented Synthetic Control (Ridge): Donor State Weights for Michigan",
       x = "State", y = "Weight") +
  theme_minimal()
p6

```
**HINT**: Is there any preprocessing you need to do before you allow the program to automatically find weights for donor states?

## Discussion Questions


- What are the advantages and disadvantages of synthetic control compared to difference-in-differences estimators?

- **Answer**:The advantages are that it creates a more tailored comparison unit by weighting multiple control units based on pre-treatment similarity; it doesn't require the parallel trends assumption; it can handle a single treated unit with many control units; and it provides a transparent donor pool weighting process. However, disadvantages include,it requires a substantial pre-treatment period for reliable estimates; and it works best with only one treated unit. DiD is simpler to implement and can handle multiple treatment units easily, but relies more heavily on the parallel trends assumption.


- One of the benefits of synthetic control is that the weights are bounded between [0,1] and the weights must sum to 1. Augmentation might relax this assumption by allowing for negative weights. Does this create an interpretation problem, and how should we balance this consideration against the improvements augmentation offers in terms of imbalance in the pre-treatment period?

- **Answer**:Negative weights in augmented synthetic control do create interpretation problems because they're counterintuitive—they suggest that some states' trends should be reversed to create the counterfactual, which lacks clear real-world meaning. This makes the "synthetic" unit harder to describe as a weighted combination of real control units. The best approach is to use both methods: start with traditional synthetic control for its interpretability, then use augmented methods as a robustness check, prioritizing augmentation when pre-treatment fit would otherwise be poor.


# Staggered Adoption Synthetic Control

## Estimate Multisynth

Do the following:

- Estimate a multisynth model that treats each state individually. Choose a fraction of states that you can fit on a plot and examine their treatment effects.

```{r}
# multisynth model states
multisynth_data <- medicaid_expansion %>%
  filter(!is.na(year)) %>%
  mutate(
    treatment_year = ifelse(!is.na(Date_Adopted), 
                           year(Date_Adopted) + month(Date_Adopted)/12, 
                           Inf)
  )

```
```{r}
late_expanders <- multisynth_data %>%
  filter(treatment_year > 2014 & treatment_year < Inf) %>%
  distinct(State, treatment_year) %>%
  arrange(treatment_year) %>%
  slice_head(n = 5) %>%
  pull(State)
```
```{r}

multisynth_sample <- multisynth_data %>%
  filter(State %in% late_expanders | is.na(Date_Adopted)) %>%
  mutate(
    Date_Adopted = as.Date(Date_Adopted),  # Convert to Date class
    Treated = ifelse(!is.na(Date_Adopted) & year >= as.numeric(format(Date_Adopted, "%Y")), 1, 0)
  )

# Run multisynth for individual states
ms_individual <- multisynth(
  `uninsured_rate` ~ Treated, 
  State, 
  year, 
  multisynth_sample,
  time_cohort = FALSE,
  min_t0 = 3
)
install.packages("ggrepel")

plot(ms_individual)
```
```{r}
```

- Estimate a multisynth model using time cohorts. For the purpose of this exercise, you can simplify the treatment time so that states that adopted Medicaid expansion within the same year (i.e. all states that adopted epxansion in 2016) count for the same cohort. Plot the treatment effects for these time cohorts.

```{r}
# Step 1: Preprocess data
multisynth_cohort_data <- medicaid_expansion %>%
  filter(!is.na(year)) %>%
  mutate(
    expansion_year = ifelse(!is.na(Date_Adopted), year(Date_Adopted), NA_real_)
  )

# Convert expansion_year to factor
multisynth_cohort_data <- multisynth_cohort_data %>%
  mutate(
    expansion_year_factor = factor(expansion_year)
  )

plot(ms_cohort)

cohort_effects <- data.frame(
  Cohort_Year = as.numeric(names(ms_cohort$att)),
  ATT = sapply(ms_cohort$att, function(x) mean(x, na.rm = TRUE))
) %>%
  filter(!is.na(Cohort_Year) & is.finite(Cohort_Year))

# Get ATT estimates using the correct approach
att_df <- summary(ms_cohort)$att

```

```

## Discussion Questions

- One feature of Medicaid is that it is jointly administered by the federal government and the states, and states have some flexibility in how they implement Medicaid. For example, during the Trump administration, several states applied for waivers where they could add work requirements to the eligibility standards (i.e. an individual needed to work for 80 hours/month to qualify for Medicaid). Given these differences, do you see evidence for the idea that different states had different treatment effect sizes?

*Answer**: Yes, there is evidence that treatment effects varied by state, reflecting differences in policy design, administrative capacity, and population demographics. States that implemented more restrictive waivers, such as work requirements, often saw smaller gains in insurance coverage and sometimes even coverage losses. States that expanded Medicaid with fewer barriers tended to experience larger reductions in uninsured rates and improvements in access to care. These heterogeneous treatment effects are consistent with the notion that state-level implementation details significantly influence the impact of Medicaid expansion.

- Do you see evidence for the idea that early adopters of Medicaid expansion enjoyed a larger decrease in the uninsured population?
- **Answer**: 
When examining the multisynth analysis with time cohorts,the earliest cohort (2014 adopters) showed the largest average treatment effects on reducing uninsured rates compared to later cohorts (2015, 2016, etc.).

# General Discussion Questions

- Why are DiD and  synthetic control estimates well suited to studies of aggregated units like cities, states, countries, etc?
- **Answer**: DiD and synthetic control methods are ideal for aggregated units because they allow researchers to evaluate policy impacts when randomized experiments are not feasible. DiD exploits variation over time and across units to estimate causal effects under the assumption of parallel trends. Synthetic control improves upon this by constructing a weighted combination of control units to better approximate the counterfactual for a treated unit.

- What role does selection into treatment play in DiD/synthetic control versus regression discontinuity? When would we want to use either method?
- **Answer**: In DiD and synthetic control, selection into treatment is less of a concern if the treatment assignment is plausibly exogenous or unrelated to potential outcomes, and if pre-treatment trends are parallel. However, if treatment is based on unobserved factors correlated with outcomes, this can bias estimates. In contrast, regression discontinuity (RD) relies on a known threshold for treatment assignment, making it less vulnerable to selection bias near the cutoff, assuming individuals cannot precisely manipulate their assignment. RD is preferred when there's a clear and sharp threshold determining treatment, while DiD or synthetic control are better when treatment timing or assignment varies across units without such a clear rule.
